---
title: "Assignment 10: Data Scraping"
author: "Tori Newton"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1: Load packages
library(tidyverse)
library(rvest)
library(lubridate)
library(here)

#Check working directory
getwd()

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2:Fetch the web resources from the URL
webpage <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024')
webpage

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3: Scraping the data
water_system_name <- webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
water_system_name

pwsid <- webpage %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text()
pwsid

ownership <- webpage %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text()
ownership

maximum_day_use <- webpage %>%
  html_nodes("th~ td+ td") %>%
  html_text()
maximum_day_use

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4: Convert scraped data into a dataframe 
months <- c("Jan", "May", "Sept", "Feb", "June", "Oct", "Mar", "July", "Nov", "April", "Aug", "Dec")

df <- data.frame(
  "Month" = months,
  "Year" = rep(2024,12),
  "Max_Day_Use" = as.numeric(maximum_day_use)) %>%
    mutate(
    Water_System_Name = rep(water_system_name, 12),
    PWSID = rep(pwsid, 12), 
    Ownership = rep(ownership, 12),
    Date = my(paste(Month, "-", Year)))
print(df)

#5: Line plot 
df %>%
  ggplot(aes(x = Date, y = Max_Day_Use)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +
  scale_x_date(
    breaks = seq(from = as.Date("2024-01-01"), to = as.Date("2024-12-01"), by = "month"),
    labels = format(seq(from = as.Date("2024-01-01"), to = as.Date("2024-12-01"), by = "month"), "%b") 
  ) +
  labs(
    title = "Maximum Daily Withdrawals Across Months for 2024",
    x = "Month",
    y = "Maximum Daily Use (MGD)")

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two input - "PWSID" and "year" - that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6: Create scraping function
scrape_data <- function(PWSID, the_year) {
#Retrieve the website contents 
  the_website <- 
    read_html(paste0
              ('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', PWSID, "&year=", the_year))
  #Scrape the data items
  water_system_name <- the_website %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
pwsid <- the_website %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text()
ownership <- the_website %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text()
maximum_day_use <- the_website %>%
  html_nodes("th~ td+ td") %>%
  html_text()
#Convert to a dataframe
df_new <- data.frame(
  "Month" = months,
  "Year" = rep(the_year,12),
  "Max_Day_Use" = as.numeric(maximum_day_use)) %>%
    mutate(Water_System_Name = rep(water_system_name, 12),
    PWSID = rep(pwsid, 12), 
    Ownership = rep(ownership, 12),
    Date = ymd(paste(Year, Month, "01", sep = "-")))
#Return the dataframe
 return(df_new)
}
```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020

```{r fetch.and.plot.Durham.2020.data}
#7: Scrape data for 2020
df_2020 <- scrape_data(PWSID= "03-32-010", the_year=2020)

#Line plot
df_2020 %>%
  ggplot(aes(x = Date, y = Max_Day_Use)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +
  scale_x_date(
    breaks = seq(from = as.Date("2020-01-01"), to = as.Date("2020-12-01"), by = "month"),
    labels = format(seq(from = as.Date("2020-01-01"), to = as.Date("2020-12-01"), by = "month"), "%b")  
  ) +
  labs(
    title = "Maximum Daily Withdrawals Across Months for 2020",
    x = "Month",
    y = "Maximum Daily Use (MGD)")
```

8. Use the function above to extract data for Asheville (PWSID = '01-11-010') in 2020. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2020.data}
#8: Scrape data for Asheville in 2020
df_asheville_2020 <- scrape_data(PWSID = "01-11-010", the_year = 2020)

#Combine data for Durham and Asheville
df_combined <- bind_rows(
  df_2020 %>% mutate(Municipality = "Durham"),
  df_asheville_2020 %>% mutate(Municipality = "Asheville")
)

#Plot comparing Asheville and Durham's water withdrawals
df_combined %>%
  ggplot(aes(x = Date, y = Max_Day_Use, color = Municipality)) +
  geom_line() +
  geom_point() +
  scale_x_date(
    breaks = seq(from = as.Date("2020-01-01"), to = as.Date("2020-12-01"), by = "month"),
    labels = format(seq(from = as.Date("2020-01-01"), to = as.Date("2020-12-01"), by = "month"), "%b")
  ) +
  labs(
    title = "Comparison of Maximum Daily Withdrawals: Asheville vs Durham (2020)",
    x = "Month",
    y = "Maximum Daily Use (MGD)"
  )
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2018 thru 2023.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Asheville.multiyear.data }
#9: Asheville 2018-2023
Asheville_PWSID <- "01-11-010"
the_years <- 2018:2023

asheville_list <- rep(Asheville_PWSID, length(the_years))

df_asheville_2018_2023 <- map2(asheville_list, the_years, scrape_data) %>%
  bind_rows()
print(df_asheville_2018_2023)

#Plot of Asheville's max daily withdrawal for 2018-2023
df_asheville_2018_2023 %>%
  ggplot(aes(x=Date, y=Max_Day_Use)) +
  geom_line() +
  geom_point() +
  geom_smooth(method = "loess") +
  scale_x_date(
    breaks = seq(from = as.Date("2018-01-01"), to = as.Date("2023-12-01"), by = "year"),
    labels = format(seq(from = as.Date("2018-01-01"), to = as.Date("2023-12-01"), 
                        by = "year"), "%b %Y")
  ) +
  labs(
    title = "Asheville's Maximum Daily Withdrawals by Month (2018-2023)",
    x = "Month",
    y = "Maximum Daily Use (MGD)",
    color = "Year"
  )

```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer: From just looking at the plot, it appears that overall Asheville has increased water usage over time. Although, the trend does fluctuate up and down with a dip in 2021 but overall the water usage in 2023 is much higher than in 2018.
>
