---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Tori Newton"
date: "Spring 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
#Check working directory
getwd()

#Load libraries
library(tidyverse)
library(agricolae)
library(lubridate)
library(here)
library(corrplot)
library(dplyr)

#Import the raw NTL-LTER raw data file
NTL_LTER <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"),
                     stringsAsFactors = TRUE)

#Set date columns to date objects
NTL_LTER$sampledate <- as.Date(NTL_LTER$sampledate, format = "%m/%d/%Y")
class(NTL_LTER$sampledate) #Now Date

#2: Built a ggplot theme and set it as default theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: The mean temperature is the same across different depths for all lakes. 
Ha: The mean temperature differs across different depths for at least one lake.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4: Wrangle dataset
tidy_dataset <- NTL_LTER %>%
  filter(format(sampledate, "%m") == "07") %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  na.omit()

#5: Scatterplot
tidy_dataset %>%
  ggplot(aes(x=depth, y=temperature_C)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue") +
  scale_y_continuous(limits = c(0,35)) +
  labs(title = "Temperature vs. Depth",
       x = "Depth (m)", y = "Temperature (°C)")

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: This plot shows a negative relationship between depth and temperature.
As depth increases, the temperature tends to decrease. This is typical in 
aquatic environments. Since the line for the linear model slopes downward, it 
confirms that there is a negative linear trend between depth and temperature. 
While the linear regression line suggests an overall negative linear 
relationship, the scatter of points indicates variability that is not perfectly 
captured by a linear model. This variability suggests that a more complex
relationship might exist at certain depths.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7: linear regression
linear_model <- lm(data=tidy_dataset, temperature_C ~ depth)
summary(linear_model)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:The residuals suggest that while the model generally provides good 
predictions, there are some larger deviations such as the min of -9.5173 and 
the max of 13.5834. These indicate that the model does not perfectly fit the 
data. The depth of -1.94621 indicates that for each 1 meter increase in depth, 
the temperature decreases by approximately 1.95°C. The negative sign confirms 
the inverse relationship between depth and temperature. Both the intercept and
depth coefficients have p-vales <2e-16 which is much smaller than 0.05, 
indicating that both are statistically significant predictors of temperature. 
The r-squared value of 0.7387 means that the model explains about 73.87% of the 
variability in temperature which is a relatively strong fit. This indicates 
that changes in depth explain a significant protion of the variability in 
temperature. The model includes only one predictor variable, which is depth. 
Therefore, the model has 1 degree of freedom for the predictor variable. For 
the f-statistics test, the very low p-value of <2.2e-16 indicates that the 
model is highly significant and that depth is a valid preditor of temperature. 

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9: run an AIC
full_model <- lm(data = tidy_dataset, temperature_C ~ year4 + daynum + depth)

best_model <- step(full_model)
summary(best_model)

#10: run a multiple regression
multiple_regression_model <- 
  lm(data = tidy_dataset, temperature_C ~ year4 + daynum + depth)
summary(multiple_regression_model)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: Based on the regression results, the final set of explanatory 
variables to predict temperature are year4, daynum, and depth. The multiple 
r-squared value is 0.7412 and the adjusted r-squared value is 0.7411. This
means that the model explains 74.12% of the variability in lake temperature
based on the predictors year4, daynum, and depth. When only depth was used as 
the explanatory varibale the r-squared value was 0.7387. This is 0.0025 or 
0.25% lower than the new r-squared value. This indicates that a 0.25% increase 
in the amount of variance was explained by adding year4 and daynum to the 
model. The inclusion of year4 and daynum improves the model only marginally. 

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12: Format ANOVA as aov
tidy.dataset.anova <- aov(data= tidy_dataset, temperature_C ~ lakename)
summary(tidy.dataset.anova)

#Format ANOVA as lm
tidy.dataset.anova2 <- lm(data= tidy_dataset, temperature_C ~ lakename)
summary(tidy.dataset.anova2)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, there is a significant difference in the mean temperature among 
the lakes based on the results of the ANOVA. The p-value of <2e-16 is much
smaller than the typical significance level of 0.05. A p-value this small 
indicates that the null hypothesis can be rejected and it can be concluded that
there is a statistically significant difference in the mean temperature across
the lakes. 

14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14: Graph of temperature by depth
tidy_dataset %>%
  ggplot(aes(x=depth, y=temperature_C, color = lakename)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_y_continuous(limits = c(0,35)) +
  labs(title = "Temperature by Depth for Each Lake",
       x = "Depth (m)", y = "Temperature (°C)", color = "Lake")

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15: Tukey test
TukeyHSD(tidy.dataset.anova)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:Paul Lake and Ward Lake have the same mean temperature as Peter Lake 
statistically speaking. If the p adj value is greater than 0.05, this indicates
that there is no significant difference. Peter Lake and Paul Lake have a p adj
value of 0.2241586, which is greater than 0.05, indicating no significant 
difference in their mean temperatures. Peter Lake and Ward Lake have a p adj 
value of 0.7827037, which is also greater than 0.05. To find a lake that is 
statistically distinct from all other lakes, the p adj value would have to be
less than 0.05 with all other lakes. None of the lakes fit this criteria, 
therefore no lake has a mean temperature that is statistically distinct from 
all the other lakes.
 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: You could use an independent t-test to compare Peter Lake and Paul
Lake. The t-test would assess whether the means of the two lakes are 
significantly different from each other. 

18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
#Filter the data to include only Crampton Lake and Ward Lake
subset_data <- tidy_dataset %>%
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))

#Run the two-sample t-test
t_test <- t.test(data = subset_data, temperature_C ~ lakename)
print(t_test)
```

>Answer: The p-value of 0.2649 is greater than 0.05, which means we fail to 
reject the null hypothesis and that mean temperatures of Crampton Lake and Ward 
Lake are equal.The mean temperature for Crampton Lake is 15.35°C and the mean 
temperature of Ward Lake is 14.46°C. While the means are different, the 
difference is small and the statistical test does not suggest that this 
difference is significant. Based on the t-test we do not have enough evidence
to conclude that  the mean temperatures of Crampton Lake and Ward Lake are 
different. From question 16, the p adj value for Ward Lake-Crampton Lake is
0.9714459 indicating no significant difference in their mean temperatures. This 
supports these findings that Crampton Lake and Ward Lake have the same mean
temperature, statistically speaking.
